Abstractions for Multi-Sorted Substitutions

### 2023-08-22 paper/paper.tex
Update.

  \subsection{Unifying Renamings \& Substitutions}
  To avoid the duplication between renamings and substitutions,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced an abstraction called \emph{Kits}.

  % While Kits were originally formulated for intrinsically-typed languages,
  % they nicely extend to our extrinsically-typed multi-sorted syntax, whose
  % sort-indexing is a form of intrinsic typing.

  A kit is a structure that abstracts over the difference between variables and terms.
  There are always exactly two instances for Kits, one for variables and one for terms,

  The core observation leading to this abstraction is that renamings
  \texttt{S₁ →ᵣ S₂}
  and substitutions \texttt{S₁ →ₛ S₂} only differ in their return type:
  while a substitution returns a term \texttt{S₂ ⊢ s}, a renaming returns
  a variable \texttt{S₂ ∋ s}.

  A Kit is a structure that abstracts precisely over the commonalities
  between variables and terms.

  % \ExampleCode\FExampleSubRen

  In both cases \texttt{\_∋\_} and \texttt{\_⊢\_} are instanciated to the type
  \LibCode\KScoped
  which allows us to create a structure to axiomatize the common
  theory behind both of them:
  \LibCode\KKit

  \subsection{Library}
  \LibCode\KAp
  \LibCode\KExt
  \LibCode\KLift
  \LibCode\KId
  \LibCode\KSingle
  \LibCode\KWeaken
  \LibCode\KEq
  \LibCode\KFunExt
  \LibCode\KIdLift
  % \LibCode\KIdLiftProof
  \LibCode\KKitNotation
  \LibCode\KTraversal
  \LibCode\KKitInstances
  \LibCode\KKitOpen
  \subsection{System F}
  \AppCode\FTraversalOp
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

@@

>>>>

  \subsection{Unifying Renamings \& Substitutions}
  To avoid the duplication between renamings and substitutions,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced an abstraction called \emph{kits}.\footnote{
    While kits were originally formulated for intrinsically-typed languages,
    they nicely extend to our extrinsically-typed multi-sorted syntax, whose
    sort-indexing is a form of intrinsic typing.
  }

  % A \texttt{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \texttt{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instanciate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

  As we intend to have exactly two kit instances, we choose names of
  the form \texttt{x/y}, where \texttt{x} is the name we choose for the
  variable instance, and \texttt{y} is the name we choose for the term
  instance.
  For example the parameter type \texttt{\_∋/⊢\_} will be instanciated
  to \texttt{\_∋\_} for the variable kit, and to \texttt{\_⊢\_} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \texttt{id/`} converts a variable \texttt{S ∋ s} into a
    \texttt{S ∋/⊢ s}. For the variable kit, \texttt{S ∋/⊢ s} is instanciated to \texttt{S ∋ s}, so this
    operation is the identity. For the term kit, \texttt{S ∋/⊢ s} is
    instanciated to \texttt{S ⊢ s}, so this operation is the
    variable constructor \texttt{`\_}.
  \item 
    \texttt{`/id} converts a \texttt{S ∋/⊢ s} into a term \texttt{S ⊢ s}
    and is dual to the \texttt{id/`} operation.
  \item 
    \texttt{wk} shifts the DeBruin-Indices in a \texttt{S ∋/⊢ s}.
    The new, unused variable \texttt{zero} can assume any sort \texttt{s'}.
    For variables \texttt{wk} is the successor \texttt{suc}.
    For terms \texttt{wk} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \texttt{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \texttt{`\_}.
  \item \texttt{`/id-injective} and \texttt{id/`-injective} state that
    \texttt{`/id} and \texttt{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \texttt{wk-id/`} characterizes the behaviour of the \texttt{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \texttt{S₁ →ᵣ S₂} and
    substitutions \texttt{S₁ →ₛ S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \texttt{S₁ →ₖ S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \texttt{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \texttt{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \texttt{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \texttt{wk}:
    \LibCode\KWeaken
  \end{itemize}

  We also specify additional notation, which makes it less cumbersome
  to explicitly pass a particular Kit to those definitions:
  \LibCode\KKitNotation

  The only task left now is to create the two instances for the kits
  and define how a map can be applied to a term.
  While the variable kit can be defined as expected, the term kit
  raises a dependency problem: The \texttt{wk} operation for terms requires us
  to already know how to apply a renaming to term.
  To solve this problem, McBride came up with a clever solution:
  Applying a map to a term, can be defined without even having any Kit-instances yet.
  Afterwards we can first define the variable kit, and in the
  definition of the term kit, we are then already allowed to apply renamings.
  After the definition of the term-kit we are then allowed to apply
  both renamings and substitutions as both kits defined.

  \LibCode\KKitVar

  \subsection{Library}
  \LibCode\KKitOpenInst
  \LibCode\KEq
  \LibCode\KFunExt
  \LibCode\KIdLift
  % \LibCode\KIdLiftProof
  \LibCode\KTraversal
  \LibCode\KKitVar
  \LibCode\KKitTerm
  \LibCode\KKitOpen
  \subsection{System F}
  \AppCode\FTraversalOp
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

@@

  \section{Beyond System F}
  \begin{itemize}
  \item Binders with multiple variables are supported and even pattern matching can be encoded.
  \item Heterogenous equality
  \end{itemize}
